\chapter{Introduction}
\section{Motivation}

Before I begin, I ask you, reader, to consider my abbreviated description of my experience of mathematics education: I sit in a classroom full of students, all who are told to practice and memorize the materials presented by the teacher. I have questions about why a formula looks the way it does. My teacher does not answer; he thinks it's not relevant to the class. In college, I again find myself sitting in a classroom full of students, my questions left again unanswered ``in the interest of time.'' I wonder sometimes whether the lecturer knows my name or even cares if I show up at all. I go on to take an advanced toplogy course. There are three women, including me, and at least ten men, in the first class. By the end of the semester, I'm the only woman left, with at least ten other men.

Mathematics has always been a difficult subject for me. In large part, it is difficult because of how difficult it is to stay awake during lectures. Perhaps laughable, but it's true; I often joke that had it not been for my persistent parents' efforts and lifelong interest in the subject, I would have easily been an art major, which requires a lot more moving of body parts and far fewer listening to lectures. In my college math lectures, professors cover proofs of some theorems and write down some definitions. If at any point I were to doze or be distracted, I would have a harder time following for the rest of the lecture, trying to play catch-up from one point to the next. If I don't make the necessary connections and absorb the important concepts right then and there, I feel behind and lost. In worries that I'll fall even further behind, I dare not look around to see how my peers are doing, or to question the validity of what I'm being taught.

In stark constrast, I've never fallen asleep or felt like I was behind in classes that are participation-based or seminar-based. Even the way students sit in the class is different. As opposed to rows and columns facing the board, desks are arranged in rough circles that have everyone face one another, so as to encourage voices from all directions, not one.

I am therefore as baffled as I am frustrated with how traditional lecture-based classes seem to prevent effective learning. Forgetting for a moment that, personally, lectures are simply difficult to be interesting for over an hour, I, a woman and student of color, rarely see a figure for me to look up to or relate to. I have sat in numerous math lectures being the sole woman in the room, questioning where all of my fellow female math majors could have gone. I have shuffled through mathematics texts wondering when the last time I read a textbook from a female author of color was, never mind see a theorem named after one. Being Asian-American, I constantly question and face biases and stereotypes of Asian-Americans, many which make me take a second glance at my pride and love of my ethnic identity.

I find that this system of learning which disregards the students in the picture is highly ineffective. The social, historical, and economical contexts in which mathematics takes part in for people especially like myself is simply non-negligible in fostering a good learning environment.

This thesis is my attempt to bring some of the problems I see into light, as a way to expose the flaws and changes necessary in mathematics pedagogy, particularly in postsecondary education. Afterwards, I look into self-regulatory methods in teaching, and in particular, I study self-paced assessments as a tool for equity in mathematics education. Using a case study done at my own institution to see the outcomes of self-paced assessment, I raise some points of discussion about future applications.

\section{Background}
\subsection{The status quo.}

The classroom setting that I described in the introduction is an example of the ``banking model'' of education, named by Paulo Freire in {\it Pedagogy of the Oppressed} in 1968. Freire draws a metaphor between this style of education to a bank, where the teacher is the depositor and the students are depositories (Friere, 1968). The teacher's tasks are to ``fill'' the students of information, and the student's role is to simply accept this information, with no particular requirement to digest its contents further or apply additional context. Thus, under this model, a good teacher is one who can give as much as they can to as many students, and a good student is one who receives and regurgitates the most with precision.

Of the various problems Freire points out about this model, he particularly emphasizes the notion that the banking model transforms students into objects that merely act as containers, devoid of critical or creative thought.

Friere is certainly not the only individual questioning the method of education utilized in classrooms. Educator and psychologist Donald Bligh presents in his book, {\it What's the Use of Lectures?}, rationale for why traditional lecture style classrooms are ineffective and outdated, supported by an exhaustive collection of studies (Bligh, 1998). His ideas are not new either, compiling hundreds of studies conducted on this topic and relating theories from many other educators. Most notably, he cites the work of Benjamin Bloom, also a professor known for proposing and driving experiments on mastery-based learning (Bligh, 1998).

Despite the existing literature about the need for improvement in the traditional systems, change has been slow going, in large part due to the difficulty of bringing changes into fruition. Take for instance educator Bob Moses, who saw struggling students in mathematics and created a nontraditional way of teaching algebra, called the Algebra Project. In his book, he describes the experience in spreading this idea across middle schools in Boston as an ``uphill slough'' (Moses, Cobb, 2001). He was turned down by principals for reasons ranging from teachers claiming that student skills not being up to par to stating that it would be too difficult to transition from the traditional approach.

Moses' experience goes to show that there exists resistance towards change, and isn't uncommon in middle school classrooms. In many respects, postsecondary mathematics in America has not changed since a century ago. For instance, white male instructors still teach lecture-based courses with weekly homework assignments and two or so exams.

\subsection{Pedagogical methods}
Throughout the 20th century, a variety of pedagogical methods came into trial across subject areas and institutions. Specifically in regards to postsecondary mathematics education, methods that have been explored include the following.

\subsubsection{Flipped classrooms}
This method of instruction involving an inversion or a ``flip'' in the classroom is best described as placing the students in a setting where the lectures are given outside of the classroom, and activities meant to be more meaningful for the learning experience take place inside the classroom (Zappe, et al., 2009). These activities, ranging from group work to solving tutorials to leading workshops, are often described as active learning and have been shown to significantly improve student performance in science and mathematics (Freeman, et al., 2014). There are ongoing studies on the effects of flipped classroom being done in undergraduate engineering and mathematics settings (Yong, et al., 2015).

\subsubsection{Inquiry-based learning}
Inquiry-based learning (IBL) describes a broad area of educational methods that focuses on student-driven instruction in the classroom. Most notably, IBL looks into how students can learn actively through questioning and creating arguments independently (Yoshinobu, 2013). Though IBL has been experimented on and researched at many different institutions for varying disciplines, it has also been involved in improving post-secondary mathematics.

One way IBL is used in postsecondary mathematics is the Moore Method, or Modified Moore Methods. In short, the Moore Method looks to have students prove theorems from scratch with minimal instructor intervention (Parker, 1992). Created by Robert Lee Moore, this method saw fallbacks in courses other than introductory topology (which is where the method was used initially), and Moore himself received a lot of criticism for forbidding African-American students to take his classes and holding misogynistic and anti-semitic viewpoints, perhaps suggesting that his method would specifically benefit white men (Ross, 2007). Modified Moore Methods were developed and introduced in an effort to reduce some of these points of improvement (Cohen, 1982).

In 2014, a study conducted across multiple institutions with a focus to evaluate the benefits of IBL gave more insight into how students ultimately fared due to IBL (Laursen, et al., 2014). Results showed that IBL removed previously seen gender gaps that disfavored women. More recently, IBL in postsecondary mathematics have recently received attention as they appeared in discussions and examples presented in an issue of {\it Problems, Resources, and Issues in Mathematics undergraduate Studies} (Katz, 2017). Various examples that appear in this issue of IBL studies, which were run on various student bodies and subjects, bring up notable questions and results.

\subsubsection{A focus on self-regulation}
Educator and psychologist Benjamin Bloom developed mastery-based learning, which expects students to have complete or near-complete mastery of concepts before proceeding further into the material (Bloom, 1968). Over the years mastery learning has taken on many variations, some more successful than others, but all focus on individual pacing and developing autonomy in a student's ability to learn. The biggest takeaway from studies done in mastery learning is the positive impact it has on students despite how much it differs from traditional methods of teaching (Zollinger, 2017, Bradley, 2017).

Self-regulation is not a method but a vital piece of mastery-based learning that asks students to identify and understand their own progression in learning. A focus that requires and trains meta-cognitive skills in learning mathematics can be useful in enabling students to grow independence and attain ownership of their own learning. Furthermore, rather than introducing large stylistic changes to the classroom, bringing in a focus of self-regulation can be relatively easily accomplished compared to flipped classrooms or IBL (Montague, 2007). While self-regulation in mathematics learning has not been explored as explicitly as the other two methods, elements of self-regulation are present and can be found.

Exactly how incorporating the ideas of self-regulation is possible is detailed  in a later chapter. In specifics, this thesis proposes a possible way to change classroom dynamics using self-paced assessments, which attempts to alter one aspect of mathematics classes for a more effective and less stressful learning experience using self-regulative practices.

\subsection{Inequity and Underrepresentation}
Postsecondary mathematics also faces issues of underrepresentation in gender and race, in particular shown by the percentage of degrees awarded in mathematics. Below are tables (\ref{table:gender}, \ref{table:race}) with data from the most recent 2015 survey conducted by the Conference Board of Mathematical Sciences (Blair, Kirkman, Maxwell, 2018).
\begin{table}[!htb]
  \begin{center}
    \begin{tabular}{c | c | c | c}
      Department & Men & Women & Total\\ \hline
      University (PhD) & 3431 (68\%) & 1645 (32\%) & 5076\\
      University (MA) & 1436 (51\%) & 1365 (49\%) & 2801\\
      College (BA) & 2529 (51\%) & 2388 (49\%) & 4917
    \end{tabular}
  \end{center}
  \caption{Number of degrees awarded in all mathematics majors, categorized by gender and department during July of 2014 to June of 2015.}
  \label{table:gender}
\end{table}

\begin{table}[!htb]
  \begin{center}
    \begin{tabular}{l | l | l | l | l | p{1.2cm} | l}
      Department & Asian & Black & Hispanic & White & AIAN/ NHPI* & Unknown\\ \hline
      University (PhD)&&&&&\\
      Men & 15 & 1 & 3 & 55 & 0 & 2\\
      Women & 5 & 0 & 1 & 16 & 0 & 1\\ \hline
      University (MA)&&&&&\\
      Men & 11 & 2 & 3 & 46 & 0 & 2\\
      Women & 6 & 1 & 1 & 26 & 0 & 1\\ \hline
      University (BA)&&&&&\\
      Men & 6 & 2 & 1 & 53 & 0 & 2\\
      Women & 4 & 1 & 1 & 30 & 0 & 1
    \end{tabular}
  \end{center}
  {\footnotesize *American Indians and Alaskan Natives / Native Hawaiians and other Pacific Islanders}
  \caption{Percentages of full time faculty members belonging to various ethnic groups by gender and by department as of fall 2015.}
  \label{table:race}
\end{table}

What is most notable from the number of degrees awarded is the sharp decline in the percentage of women receiving PhDs than men compared to the percentage of MAs and BAs awarded. Table \ref{table:race} further shows there exist fewer women, regardless of ethnicity, in tenured faculty positions overall which diverge even more from the numbers of PhDs awarded. Underrepresentation in science and engineering as described by the National Science Foundation refers to a minority group whose number of scientists and engineers per 10,000 population of that group is substantially below the comparable figure for scientists and engineers who are white and not of Hispanic origin (United States Code, 2011). The numbers above show an example of underrepresentation of women in mathematics, a phenomenon that is important to take note of, but describing the reasons why such is the case is outside the scope of this thesis.

Looking specifically at ethnic groups, the numbers show an overall evidence of minoritization of non-white faculty for both genders. That the presented data only respects the notion of binary genders also points to a possible lack of support and representation of genderqueer or nonbinary individuals. How this minoritization and underrepresentation could potentially impact mathematics education is explored in the next chapter where I discuss inequitable practices in mathematics education.

\section{Self-paced assessment}

There is no concrete, definite solution to the problem of how mathematics education has been unfair, benefitting only some and hence being ineffective for others. Yet, it is important to try to attempt and propose possible ways to alleviate and work towards what could better serve the current student population.

One attempt to do so is through implementing self-paced assessment, a method that takes from self-regulation and self-regulatory methods. Self-paced assessment takes elements from some non-traditional methods in postsecondary mathematics education that have been experimented on previously. Such include but aren't limited to mastery-based learning, flipped classrooms, and inquiry-based learning.

In describing how self-paced assessment can be run in a postescondary mathematics classroom, this thesis presents a case study conducted on a group of first year students taking an introductory linear algebra course at Harvey Mudd College, a small liberal arts college located east of Los Angeles. The results of the study propose a possible question of using self-paced assessments to take a step towards equity by trying to shift the locus of control to students and providing a more positive experience to reduce negative sentiments towards mathematics.
